### 2015-05-11 papers/histogram/thesis/thesis.tex
fixed some mistakes in figure labels; changed sample density of states figure in thesis; inserted a small joke in my thesis

Figure \ref{fig:dos_sample} provides an example of an unnormalized
density of states $\tilde D\p{E}$ in energy for a particular
square-well fluid. The density of states has been ``normalized'' to
have a maximum value of $1$ in order to emphasize the logarithmic span
of $D\p{E}$. This logarithmic span ($\sim10^{60}$) is proportional to
the number of moves which one must simulate via Metropolis Monte Carlo
in order to observe the energies with the lowest state densities. For
reference, the world's current fastest supercomputers can perform
$\sim10^{16}$ floating-point operations per second.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-example.pdf}
  \caption[Density of states]
  {Sample density of states for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.3$, and $N=25$ spheres, as
    computed by several histogram methods. The density of states has
    been normalized to have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

Histogram methods provide a means to address Metropolis Monte Carlo's
inability to sufficiently sample low entropy states by introducing a
bias into the otherwise random sampling of state space. A weighting
function $w$ is introduced, whose domain is the value of some property
$X\p{s}$ which depends on the system state $s$. An additional
condition is then added to step \ref{alg:metropolis_move} of Algorithm
\ref{alg:metropolis} in order to accept an attempted move: the weights
$w\p{X_i}$ and $w\p{X_f}$ of the initial (pre-move) state $s_i$ and
final (post-move) state $s_f$, respectively, are used to determine the
probability $P_m\p{s_i\to s_f}$ of accepting an otherwise valid move
via
\begin{align}
  P_m\p{s_i\to s_f}=\min\set{\f{w\p{X_f}}{w\p{X_i}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\p{X_f}>w\p{X_i}$, the move $s_i\to
s_f$ is accepted; when $w\p{X_f}<w\p{X_i}$, the ratio of these weights
determines the probability of accepting the move. Due to the fact that
only ratios of weights determine $P_m\p{s_i\to s_f}$, the weights
$w\p{X}$ are scale-invariant, meaning that their effect on simulations
is unchanged by scale factors that are constant with respect to
$X\p{s}$. Furthermore, simulating with flat weights $w\p{X}=w_0$ for
all $X$ is equivalent to simulating without weights, as the
probability $P_m\p{s_i\to s_f}$ will always equal 1.

@@

Figure \ref{fig:transitions_sample} provides an example of a
transition matrix $T_d\p{E,\Delta E}$ for a particular square-well
fluid. The band at $\Delta E=0$ indicates that most moves transition
from an energy to itself. Low energies are more likely to transition
up in energy, with $\Delta E>0$, than down in energy, with $\Delta
E<0$. The high energies which are more likely to transition down in
energy than up are those above the state of maximum entropy.

@@

I wish to thank, first and foremost, my advisor Dr. David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
codes. Lastly, I would like to thank thank Eva Morgun, who prompted me
to {\it act} on my textbook enthusiasm for physics, and get involved
in the fun and interesting world of scientific research.

>>>>

Figure \ref{fig:dos_sample} provides an example of an unnormalized
density of states $\tilde D\p{E}$ in energy for a particular
square-well fluid. The density of states has been ``normalized'' to
have a maximum value of $1$ in order to emphasize the logarithmic span
of $D\p{E}$. This logarithmic span ($\sim10^{300}$) is proportional to
the number of moves which one must simulate via Metropolis Monte Carlo
in order to observe the energies with the lowest state densities. For
reference, there are approximately $10^{80}$ atoms in the observable
universe, and the world's current fastest supercomputers can perform
about $10^{16}$ floating-point operations per second.

\begin{figure}
  \centering
  \includegraphics[width=0.6\textwidth]{figs/dos-example.pdf}
  \caption[Density of states]
  {Sample density of states for square-well fluid with a well width
    $\lambda=1.3$, filling fraction $\eta=0.1$, and $N=100$ spheres,
    computed via the transition matrix Monte Carlo method. The density
    of states has been normalized to have a maximum value of $1$.}
  \label{fig:dos_sample}
\end{figure}

@@

We do not have time to count all of the atoms in the observable
universe, and we certainly do not have time to wait for Metropolis
Monte Carlo to sample (and worse, collect decent statistics on) low
energy square-well fluid states. Histogram methods provide a means to
address Metropolis Monte Carlo's inability to sufficiently sample low
energy states by introducing a bias into the otherwise random
sampling of state space. A weighting function $w$ is introduced, whose
domain is the value of some property $X\p{s}$ which depends on the
system state $s$. An additional condition is then added to step
\ref{alg:metropolis_move} of Algorithm \ref{alg:metropolis} in order
to accept an attempted move: the weights $w\p{X_i}$ and $w\p{X_f}$ of
the initial (pre-move) state $s_i$ and final (post-move) state $s_f$,
respectively, are used to determine the probability $P_m\p{s_i\to
  s_f}$ of accepting an otherwise valid move via
\begin{align}
  P_m\p{s_i\to s_f}=\min\set{\f{w\p{X_f}}{w\p{X_i}},1}.
  \label{eq:move_prob}
\end{align}
This formula means that when $w\p{X_f}>w\p{X_i}$, the move $s_i\to
s_f$ is accepted; when $w\p{X_f}<w\p{X_i}$, the ratio of these weights
determines the probability of accepting the move. Due to the fact that
only ratios of weights determine $P_m\p{s_i\to s_f}$, the weights
$w\p{X}$ are scale-invariant, meaning that their effect on simulations
is unchanged by scale factors that are constant with respect to
$X\p{s}$. Furthermore, simulating with flat weights $w\p{X}=w_0$ for
all $X$ is equivalent to simulating without weights, as the
probability $P_m\p{s_i\to s_f}$ will always equal 1.

@@

Figure \ref{fig:transitions_sample} provides an example of a
transition matrix $T_d\p{E,\Delta E}$ for a particular square-well
fluid. The band at $\Delta E=0$ indicates that most moves transition
from an energy to itself. Low energies are more likely to transition
up in energy, with $\Delta E>0$, than down in energy, with $\Delta
E<0$. The high energies which are more likely to transition down in
energy than up are those above the state of maximum entropy, i.e. the
state at which $D\p{E}$ is maximal.

@@

I wish to thank, first and foremost, my advisor Dr. David Roundy, for
mentoring me and providing me a wealth of knowledge and expertise in
various subjects of physics, numerical methods, and programming
nuances. I would also like thank Paho Lurie-Gregg, with whom I
frequently exchange and discuss codes and ideas, and whose hard
polyhedra Monte Carlo simulations I used as the template for my own
codes. Lastly, I would like to thank thank Eva Morgun, who prompted me
to {\it act} on my textbook enthusiasm for physics, and get involved
in the fun and interesting world of scientific research. \fixme{Thank
  Jansen and McIntyre}
